---
title: "assignment 3 -fml"
author: "Chirag gupta"
date: "2025-10-08"
output: html_document
---
```{r}
library(e1071)    # for naiveBayes
library(reshape2) # for melt and cast
library(dplyr)    # for data wrangling
library(caret)    #for classification
library(readr)    #to read csv file
```

```{r}
#reading csv file
orignal_bank <- read_csv("C:/Users/devil/Downloads/UniversalBank (2).csv")
View(orignal_bank)
names(orignal_bank)
```
```{r}
#keeping the required coloums as asked 
bank <- orignal_bank[, c("Online", "CreditCard", "Personal Loan")]
set.seed(123)
```
```{r}
colnames(bank) <- c("Online", "CC", "Loan") #remaining the coloum names
head(bank)
```
```{r}
#converting variables in factor
bank$Online <-as.factor(bank$Online)
bank$CC <- as.factor(bank$CC)
bank$Loan <- as.factor(bank$Loan)
```

```{r}
#partitoning the dataset in 60 percent train and remaining in valid 
train_index <- createDataPartition(bank$Loan, p = 0.6, list = FALSE)

train <- bank[train_index, ]
valid <- bank[-train_index, ]

# Check dimensions of training set
dim(train)

# Check dimensions of validation set
dim(valid)
```
```{r}
#Answer-1
pivot_table <- with(train, table(CC, Online,Loan))
print(pivot_table)
ft <- ftable(train$CC, train$Loan, train$Online)
ft

```

ALTERNATIVELY FOR A BETTER UNDERSTANDING 

```{r}
melt <- melt(train, id.vars = c("CC", "Online", "Loan"))
cast_table <- dcast(melt, CC + Loan ~ Online, length) 
print(cast_table)
```


INTERPRETATION - 

In the above code i have used both the functions for a clearer view of our data where cc and loan are used as row and online as coloumn, wherein the values 0 in online shows how many have not taken the cc and loan and 1 represent vice verse
```{r}
#Answer-2 Find probability of Loan = 1 given CC = 1 and Online = 1

# We will filter training data for customers with CC=1 and Online=1
subset_data <- train %>%
  filter(CC == 1, Online == 1)

# Calculate probability of loan acceptance in this group
prob <- sum(subset_data$Loan == 1) / nrow(subset_data)

cat("P(Loan = 1 | CC = 1, Online = 1) from data (pivot table) =", prob, "\n") # concatinating and printing at same time.
```
INTERPRETATION - 

In the above code it shows the  probability that a customer who owns a credit card and uses online banking accepts the loan.  IN other words- if we pick a customer who uses the bank’s credit card and online banking services, there is approximately a 9.85% chance that they will accept a personal loan
h
```{r}
#Answer-3 
#Creating two tables for training data
#Table 1: Relationship between Loan and Online
# Create the table
table_loan_online <- table(train$Loan, train$Online)

# Add row and column names
colnames(table_loan_online) <- c("Online = 0", "Online = 1")
rownames(table_loan_online) <- c("Loan = 0", "Loan = 1")

# Print nicely
print(table_loan_online)

```


```{r}
# Create the table
table_loan_cc <- table(train$Loan, train$CC)

# Add readable row and column names
rownames(table_loan_cc) <- c("Loan = 0", "Loan = 1")
colnames(table_loan_cc) <- c("CC = 0", "CC = 1")

# Print the formatted table
print(table_loan_cc)


```
INTERPRETATION- 

Int his step we are creating 2 tables - one by using loan and online and the other by using loan and cc where loan has been taken as rows in both of the tables 


```{r}
# Answer-4 Computing the probablites as asked 
# P(CC = 1 | Loan = 1)
P_CC1_L1 <- sum(train$CC == 1 & train$Loan == 1) / sum(train$Loan == 1)

# P(Online = 1 | Loan = 1)
P_Online1_L1 <- sum(train$Online == 1 & train$Loan == 1) / sum(train$Loan == 1)

# P(Loan = 1)
P_L1 <- mean(train$Loan == 1)

# P(CC = 1 | Loan = 0)
P_CC1_L0 <- sum(train$CC == 1 & train$Loan == 0) / sum(train$Loan == 0)

# P(Online = 1 | Loan = 0)
P_Online1_L0 <- sum(train$Online == 1 & train$Loan == 0) / sum(train$Loan == 0)

# P(Loan = 0)
P_L0 <- mean(train$Loan == 0)

# results
cat("P(CC=1|Loan=1):", P_CC1_L1, "\n")
cat("P(Online=1|Loan=1):", P_Online1_L1, "\n")
cat("P(Loan=1):", P_L1, "\n")
cat("P(CC=1|Loan=0):", P_CC1_L0, "\n")
cat("P(Online=1|Loan=0):", P_Online1_L0, "\n")
cat("P(Loan=0):", P_L0, "\n")
```
INTERPRETATION- 

In this step we are finding the conditional probablites for credit card holder among loan acceptors with different conditons given where cc, online, loan are 0,1

So when- 

CC=1|Loan=1        probabilty= 0.2916667 

Online=1|Loan=1    probabilty= 0.6111111 

Loan=1             probabilty=0.096 

CC=1|Loan=0        probabilty=0.2865044

Online=1|Loan=0    probabilty=0.5940265 

Loan=0             probabilty=0.904 

```{r}
#ASNWER- 5 Computing Naive Bayes probability manually

# P(Loan=1 | CC=1, Online=1) = [P(CC=1|Loan=1) * P(Online=1|Loan=1) * P(Loan=1)] /
#                              ([P(CC=1|Loan=1)*P(Online=1|Loan=1)*P(Loan=1)] + [P(CC=1|Loan=0)*P(Online=1|Loan=0)*P(Loan=0)])

num <- P_CC1_L1 * P_Online1_L1 * P_L1
den <- num + (P_CC1_L0 * P_Online1_L0 * P_L0)
P_L1_given_CC1_Online1 <- num / den

cat("Naive Bayes estimated P(Loan = 1 | CC = 1, Online = 1) =", P_L1_given_CC1_Online1, "\n")

```

INTERPRETATION - 

Here we are computing the naive bayes probablity ont he above computed entries and printing the result for loan=1,cc=1,online=1 for comparing it 


```{r}
# Asnwer-6. Comparing the two probabilities

cat("Empirical (from data)  :", prob, "\n")
cat("Naive Bayes (from model):", P_L1_given_CC1_Online1, "\n")
```
INTERPRETATION -

Naive bayes model is more accurate as  we can easily see that the probabilty we we derived from actual data and the probabilty which we derived from naive bayes are quite close but naive bayes model is more accurate since it has been predicted it can be applied to unseen data and the results will be more accurate than the emperical data if applied to unseen data. 

```{r}
# ANSWER-7 Run Naive Bayes model on data


# Train Naive Bayes model using CC and Online to predict Loan
nb_model <- naiveBayes(Loan ~ CC + Online, data = train,laplace = 1)

c <-predict(nb_model,train, type = "raw")
head(c)
```


```{r}
# Predict probability for a new customer with CC=1 and Online=1
new_customer <- data.frame(CC = factor(1, levels = c(0, 1)),
                           Online = factor(1, levels = c(0, 1)))
predicted_prob <- predict(nb_model, new_customer, type = "raw")
predicted_prob
```
INTERPRETATION- 

In this step we are running the naive bayes on the actual data and then compare with the naive bayes probabilty. Upon finding we can actually see that the model has returned the almost same answer in both the cases the slight difference is because of the laplace function otherwise it would have returned the same answer.
 

 


 
 
 
 
 
 