---
title: "Assignment 2 -fml"
author: "Chirag gupta"
date: "2025-09-26"
output: html_document
---

```{r}
## Loading necessary libraries 
library(caret) #For Knn Classification
library(class) #For data preprocessing 
library(readr) #to read csv file
```



```{r}
# QUESTION 1- Importing and Preparing universal bank data set
bank <- read_csv("C:/Users/devil/Downloads/UniversalBank.csv")
summary(bank) # To check the overall summary and get an insight
str(bank)  # To check that what are the class of our data set(which is numerical or categorical or anything else.)
```
```{r}
#removing unwanted variable id, zipcode
new_bank<-bank[, -c(1,5)]
new_bank

```
```{r}
#Converting variable(EDUCATION)
new_bank$Education <- as.factor(new_bank$Education)
str(new_bank)
names(new_bank$Education)
```


```{r}
#Conevrting categorical variable into dummy variable and getting a new data set
dummy_education <- dummyVars(~., data =new_bank)
new_bank <- as.data.frame(predict(dummy_education, new_bank))
head(new_bank)
```
```{r}
#removing the backtick from personal loan
names(new_bank) <- make.names(names(new_bank))
#removing x from them
clean_names <- function(names_vec) {
  names_vec <- gsub("^X\\.", "", names_vec)   # Remove leading 'X.'
  names_vec <- gsub("\\.$", "", names_vec)    # Remove trailing '.'
  return(names_vec)
}
names(new_bank) <- clean_names(names(new_bank))
colnames(new_bank)


```

```{r}
##Answer 1: Splitting Dataset into Training, Validation Sets
set.seed(123)
train_Index <- createDataPartition(new_bank$Personal.Loan, p=0.6, list=FALSE)
Train <- new_bank[train_Index,]
validation <- new_bank[-train_Index,]
#excluding target variable personal loan from it
train_predictor <-Train[,-10]
validation_predicator <- validation[,-10]
colnames(train_predictor)
```
```{r}
#Normalizing the train and validation
new_bank_normalized <-preProcess(Train[,-10], method=c("center", "scale"))
train_label <-predict(new_bank_normalized,Train[,-10])
validation_label <-predict(new_bank_normalized,validation[,-10])
```
Interpretation
The universal bank dataset, is divided into two sections in the section above:60% for training, 40% for validation, and later removed the target variable(personal loan) and then normalized the data.

```{r}
#ANSWER 2- k-NN Classification
#Adding a new customer
new_customer_record <- data.frame(
  Age = 40,
  Experience = 10,
  Income = 84,
  Family = 2,
  CCAvg = 2,
  Education.1 = 0,
  Education.2 = 1,
  Education.3 = 0,
  Mortgage = 0,
  Securities.Account = 0,
  CD.Account = 0,
  Online = 1,
  CreditCard = 1
)
#Normalizing new data
names(new_bank_normalized)
new_customer <-new_customer_record
new_customer <-predict(new_bank_normalized, new_customer_record)

```
Interpretation
In the above section, a data frame for the new sample dataset is created using the specified variable values, and the data frame is normalized using the same normalization parameters as the bank dataset.

```{r}
#Determining the Optimal k Value
# Running k-NN classification with k = 1 for the new customer record
knn.prediction1 <- class::knn(train = train_label, 
                       test = new_customer, 
                       cl = Train$Personal.Loan, k = 1)
knn.prediction1
```
Interpretation
The k-NN model classified the new customer entry as 0,meaning that the new customer will not accept the loan
```{r}
# ANSWER-3 Finding the value of k that balance between underfitting and over fitting 
accuracy_data <- data.frame(k = seq(1, 15, 1), overallaccuracy = rep(0, 15))
for (i in 1:15) {
  knn.pred<-class::knn(train=train_label,
                       test=validation_label,
                       cl=Train$Personal.Loan, k=i)
  
  accuracy_data[i,2]<-confusionMatrix(knn.pred,as.factor(validation$Personal.Loan),positive ="1")$overall[1]
  
}
which(accuracy_data[,2]==max(accuracy_data[,2]))
```
Interpretation :
The k-NN model was evaluated for k. loop for K values between 1 and 15 are used to determine the appropriate balance of overfitting to underfitting. which returned k=1 which means the maximum overall accuracy attained at k=1.

```{r}
# Plotting accuracy results against k values
plot(accuracy_data$k,accuracy_data$overallaccuracy)

```
```{r}
# STEP 4: Confusion matrix for validation set using best k
best_k <- 1   # from above loop
knn.pred.best <- class::knn(train = train_label,
                            test = validation_label,
                            cl = Train$Personal.Loan,
                            k = best_k)

conf_matrix <- confusionMatrix(knn.pred.best,
                               as.factor(validation$Personal.Loan),
                               positive = "1")
conf_matrix
```
INTERPERTATION 
This gives accuracy, sensitivity, specificity, and the full confusion matrix for the validation data at best k.

Sensitivity = ability to correctly predict loan acceptance (class 1).

Specificity = ability to correctly predict non-acceptance (class 0)
```{r}
# STEP 5: Classifying the new customer with best k
new_customer_pred <- class::knn(train = train_label,
                                test = new_customer,
                                cl = Train$Personal.Loan,
                                k = best_k)
new_customer_pred
```
Interpretation:
The output is 0 which means that the new customer has not accepted the loan. 
```{r}
#ANSWER- REPARTATION OF DATASET 
# STEP 6: Repartition into Train:Validation:Test (50:30:20)
set.seed(123)
train_index2 <- createDataPartition(new_bank$Personal.Loan, p = 0.5, list = FALSE)
Train2 <- new_bank[train_index2, ]
temp <- new_bank[-train_index2, ]

# split temp into validation (30%) and test (20%)
val_index <- createDataPartition(temp$Personal.Loan, p = 0.6, list = FALSE)
Validation2 <- temp[val_index, ]
Test2 <- temp[-val_index, ]

# prepare predictors (exclude target var)
train_pred2 <- Train2[, -10]
val_pred2   <- Validation2[, -10]
test_pred2  <- Test2[, -10]

# normalize with training
norm2 <- preProcess(train_pred2, method = c("center", "scale"))
train_norm2 <- predict(norm2, train_pred2)
val_norm2   <- predict(norm2, val_pred2)
test_norm2  <- predict(norm2, test_pred2)
```
```{r}
#NOW RUNNING WITH BEST VALUE OF K 
# Training prediction
train_knn <- class::knn(train = train_norm2,
                        test = train_norm2,
                        cl = Train2$Personal.Loan,
                        k = best_k)

train_conf <- confusionMatrix(train_knn, as.factor(Train2$Personal.Loan), positive = "1")
train_conf

# Validation prediction
val_knn <- class::knn(train = train_norm2,
                      test = val_norm2,
                      cl = Train2$Personal.Loan,
                      k = best_k)

val_conf <- confusionMatrix(val_knn, as.factor(Validation2$Personal.Loan), positive = "1")
val_conf

# Test prediction
test_knn <- class::knn(train = train_norm2,
                       test = test_norm2,
                       cl = Train2$Personal.Loan,
                       k = best_k)

test_conf <- confusionMatrix(test_knn, as.factor(Test2$Personal.Loan), positive = "1")
test_conf
```


INTERPERATION 
training Set:
TN (True Negatives) = 2271
FP (False Positives) = 0
FN (False Negatives) = 0
TP (True Positives) = 229
Accuracy = (TP + TN) / (TP+TN+FP+FN)
= (229 + 2271) / (229 + 2271 + 0 + 0) = 1.000 (100%)

Sensitivity = TP / (TP+FN) = 229 / (229+0) = 1.0000 (100%)
Specificity = TN / (TN+FP) = 2271 / (2271+0) = 1.0000 (100%)

The training model is perfectly fitting the training set. This is expected with k=1, but it may indicate overfitting since real-world data won’t be classified this perfectly. 

validation set:
TN = 1335
FP = 22
FN = 42
TP = 101
Accuracy = (101+1335) / (101+1335+22+42) = 1436 / 1500 = 0.9573 (95.73%)
accuracy drops slightly to 95.7%.

Sensitivity = 101 / (101+42) = 0.706 (70.63%)
Specificity = 1335 / (1335+22) = 0.9838 (98.38%)

The model is very strong at identifying non-loan customers (specificity ~98%).

However, it struggles more to detect loan acceptances (sensitivity ~71%), meaning it misses about 3 out of 10 actual positives.

test set:
TN = 885
FP = 7
FN = 29
TP = 79
Accuracy = (79+885) / (79+885+7+29) = 964 / 1000 = 0.964 (96.4%)

Sensitivity = 79 / (79+29) = 0.7315 (73.15%)
Specificity = 885 / (885+7) = 0.9922 (99.22%)

The model achieves 96.4% accuracy.

Again, specificity is very high (99.2%), so the model is excellent at ruling out non-loan customers.

Sensitivity is 73.1%, which means it correctly identifies about 3 out of 4 loan customers but still misses some.
  

```

